---
title: "Testing MAUP with an Offset Grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing MAUP with an Offset Grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(priogrid)
```

The modifiable areal unit problem (MAUP) is the observation that aggregating
the same underlying data onto different arbitrary spatial partitions can
change the aggregated results, even when nothing about the underlying data
changes. Because PRIO-GRID cell boundaries are an arbitrary 0.5-degree lattice
with no relationship to the phenomena measured within it, it is a natural
candidate for a MAUP sensitivity check: build the same variable twice, once
on the standard grid and once on a grid shifted by a small, fixed distance,
and compare.

`pg_offset_grid()` creates the shifted config for exactly this purpose.

## Shifting a Configuration

`pg_offset_grid()` takes a `pg_config`, a distance in kilometres, and a
compass bearing, and returns a new config with the same `nrow`, `ncol` and
`crs` but a translated `extent`:

```{r offset-basic}
cfg <- pg_config(
  nrow   = 100,
  ncol   = 100,
  extent = c(xmin = 0, xmax = 10, ymin = 45, ymax = 55)
)

cfg_east <- pg_offset_grid(cfg, distance_km = 5, bearing_deg = 90)

cfg$extent
cfg_east$extent
```

`bearing_deg` follows compass convention: `0` is north, `90` (the default) is
east, `180` is south, `270` is west. For a geographic CRS such as the default
`epsg:4326`, PRIOGRID converts the kilometre distance to degrees using the
extent's vertical midpoint as a reference latitude, since a degree of
longitude covers less ground away from the equator. Pass `reference_lat`
explicitly if you want the offset to be exact at a specific latitude (e.g.
the centroid of your study area) instead:

```{r offset-reference-lat}
pg_offset_grid(cfg, distance_km = 5, bearing_deg = 90, reference_lat = 50)$extent
```

Both grids have the same resolution and cell count — only their position
changes:

```{r offset-compare, eval=FALSE}
pg          <- prio_blank_grid(cfg)
pg_east     <- prio_blank_grid(cfg_east)

terra::res(pg) == terra::res(pg_east)    # TRUE: identical cell size
terra::ext(pg) == terra::ext(pg_east)    # FALSE: shifted position
```

## Building Both Grids

Every `gen_*()` variable function, and `calc_pg()`/`build_pg_dataset()` that
call them, take a `config` argument and use it to build `prio_blank_grid()`
internally. So building a MAUP comparison is just calling the same variable
twice, once per config:

```{r build-both, eval=FALSE}
pg_set_rawfolder("/path/to/your/data/folder")

cfg <- pg_config(
  nrow       = 40,
  ncol       = 40,
  extent     = c(xmin = 5, xmax = 25, ymin = 45, ymax = 55),
  start_date = as.Date("2020-12-31"),
  end_date   = as.Date("2020-12-31")
)
cfg_east <- pg_offset_grid(cfg, distance_km = 5, bearing_deg = 90)

calc_pg("ghsl_population_grid", config = cfg)
calc_pg("ghsl_population_grid", config = cfg_east)

pop_baseline <- load_pgvariable("ghsl_population_grid", config = cfg)
pop_offset   <- load_pgvariable("ghsl_population_grid", config = cfg_east)
```

Because the grids share the same `nrow`/`ncol`, cells line up positionally
even though `pgid` is config-dependent, so the two rasters can be compared
directly (e.g. `terra::global(pop_baseline - pop_offset, "sum", na.rm = TRUE)`
for the aggregate difference, or a per-cell percent-change map).

## Taming Large Downloads

The two sources most likely to trip up a first run — `GHSL GHS Population
Grid` and `HILDA+` land cover — are distributed as a handful of global files,
several gigabytes combined, regardless of how small your `config$extent` or
date range is. `automatic_download = TRUE` (the default) fetches whatever is
missing the first time a `gen_*()`/`read_*()` function needs it, which can
look like the download is "stuck" on a slow connection.

A few things make this manageable:

- **You only pay the download cost once.** The baseline and offset grid in
  the example above use the same raw source files — `pg_offset_grid()` only
  changes how they're aggregated, not what's downloaded. Run the download
  once, then experiment with as many offsets as you like.
- **Download just the sources you need, ahead of time**, rather than letting
  a `calc_pg()` call trigger it mid-computation:

  ```{r download-targeted, eval=FALSE}
  pgsearch("HILDA")$in_name$id     # find the source id
  pgsearch("GHSL Population")$in_name$id

  files <- pg_rawfiles() |>
    dplyr::filter(source_name %in% c("HILDA+", "GHSL GHS Population Grid"))

  download_pg_rawdata(file_info = files)
  ```

- **Downloads resume automatically.** If a transfer is interrupted, re-running
  `download_pg_rawdata()` (or just retrying the `gen_*()`/`calc_pg()` call
  that triggered it) picks up from where it left off rather than starting
  over — look for `.part` files in your raw folder.
- **Check what's actually missing** with `pg_data_availability()` before
  assuming a download failed outright; `n_partial` tells you if a resumable
  transfer is already most of the way there.
- If a source keeps failing outright (not just slow), increase `max_retry`
  and `retry_max_delay` in `download_pg_rawdata()` — the defaults back off up
  to 120 seconds between attempts, which can still be too fast for a
  congested link.

## Next Steps

- [Custom Spatial and Temporal Configurations](custom-config.html) — the full
  `pg_config()` reference this vignette builds on
- [Accessing PRIOGRID as Rasters](working-with-rasters.html) — comparing and
  plotting the resulting `SpatRaster` objects
